---
title: "PathWayAnalysis"
author: "Natalina Elliott"
date: '2022-10-21'
output:
  pdf_document: default
  html_document: default
---

This is the pathway analysis practical of the OBDS course 
- The first part of this practical is using gprofiler to perform ORA on output dataset of differentially expressed genes from DESeq2

# load packages 
```{r}
# load librarys

library(tidyverse)
library(gprofiler2)
```

# Read in our tables from this morning in a new Rmd file
```{r}
# read in files from this morning 

# 1. all results (this was filtered to remove non DE genes in our DESeq2 list so only contains genes expressed in our samples) -> call this table 'all_genes_df'  - this will be our background
all_genes_df <- read_csv("CD8_vs_CD4_DE_results_lfcshrinkage_allgenes.csv")

#filter out the genes which padj is NA 

all_genes_df_filtered <- all_genes_df%>%dplyr::filter(!is.na(padj))

# 2. all significant upregulated genes -> call this upreg_genes_df
upreg_genes_df <- read_csv("CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_up.csv")

# 3. all significant down regulated genes -> call this downreg_genes_df 
downreg_genes_df <- read_csv("CD8_vs_CD4_DKO_DE_results_lfcshrinkage_0.05_log2FC1_down.csv")



# Check the dimensions of each of your dataframes using dim()
dim(all_genes_df)
dim(all_genes_df_filtered)
dim(upreg_genes_df)
dim(downreg_genes_df)
# upreg_genes_df should be 915
# downreg_genes_df should be 1204
# all_genes_df_filtered should be 17729


```
Now we want a vector of gene ids to use for analysis 

```{r}
# from each of these dataframes get a vector of the ensembl ids 
all_gene_ids <- all_genes_df_filtered$ENSEMBL_geneID
    
upreg_gene_ids <- upreg_genes_df$ENSEMBL_geneID
    
downreg_gene_ids <- downreg_genes_df$ENSEMBL_geneID

all_gene_ids
```

#Perform ORA using g:profiler 

1. Look at documentation - https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
```{r}
# use the gost() function - what parameters should we include/change 
# we want to test the ORA of upreg_gene_ids vs our all_gene_ids

# get help for function
?gost() 

# write query and assign to output variable named `gost_results_obj`
gost_UPreg_obj <- gost(query=upreg_gene_ids, 
                organism = "mmusculus", ordered_query = FALSE, 
                multi_query = FALSE, significant = TRUE, exclude_iea = TRUE, 
                measure_underrepresentation = FALSE, evcodes = FALSE, 
                user_threshold = 0.05, correction_method = "g_SCS", 
                domain_scope = "annotated", custom_bg = all_gene_ids, 
                numeric_ns = "", sources = NULL, as_short_link = FALSE)
str(gost_UPreg_obj)
head(gost_UPreg_obj
# what are our outputs  


```

```{r}
# plot our results using in built function in gprofiler -> which one is it? 
gostplot(gost_UPreg_obj, capped = F, interactive = T)
#this plot is interactive and can hoover over any dot to see information
```

```{r}
# what does as_short_link = TRUE do in gost() function? 
#"returns only the web tool link to the results as a character string." making it impossible to use head or string as the outout is a weblink.  
gost(query=upreg_gene_ids, 
                organism = "mmusculus", ordered_query = FALSE, 
                multi_query = FALSE, significant = TRUE, exclude_iea = TRUE, 
                measure_underrepresentation = FALSE, evcodes = FALSE, 
                user_threshold = 0.05, correction_method = "g_SCS", 
                domain_scope = "custom", custom_bg = all_gene_ids, 
                numeric_ns = "", sources = NULL, 
     as_short_link = TRUE)

```

Answer -> "returns only the web tool link to the results as a character string." making it impossible to use head or string as the outout is a weblink.  

```{r}
# Filter results table for genesets with between 5-300 genes 



```


2. Try querying multiple gene lists at the same time (up and down regulated genes) using Multiple queries - see the documentation 

```{r}

```

Can you plot these querys together?

```{r}


```

Output the get_version_info() to save the versions of the databases that were used for the analysis!! 
```{r}
```


```{r}
#GProfiler command
get_version_info(organism ="mmusculus")
```











Message obds-22-autumn-r






```{r}
sessionInfo()

#***Do this at the end of EVERYDAY***
```


Shift + Return to add a new line
Slack needs your permission to enable notifications. Enable notifications